Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-09-17 paper/paper.Rmd
use new method categorization (in table and text)

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table&nbsp;\@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization options.
The available methods are further discussed in section \@ref(sec:disc).

\begin{table}[h]
  \centering
  \footnotesize
  \caption[t]{Spatiotemporal resampling methods implemented in the mlr3spatiotempcv package, ordered alphabetically by class name (column 'mlr3 Notation').
    Column count refers to the number of studies which were found to have used this resampling technique until May 2021.
    For each method, up to three randomly chosen references were added to the table as footnotes.}
  \begingroup
    \begin{tabular}{lllll}
      \\
      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Count\\
      \toprule
      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}}}  & \specialcell{2 \\ 3}   \\
      \cmidrule{2-5}
                                      & Spatial Blocking            & \specialcell{\pkg{blockCV}  \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
                                        & \specialcell{28 \\ 4} \\
  \cmidrule{2-5}
                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} &  21 \\
                                      \cmidrule{2-5}
       & Group-Level CV & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"custom\_cv"} \tablefootnote{\label{brenning2006}\cite{brenning2006}} \tablefootnote{\label{pena2015}\cite{pena2015}}} & \specialcell{12 \\ 2}       \\
      \midrule
      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"} \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & 6 \\
                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &  -  \\
      \midrule
      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"} \footref{morera2021}   & 1    \\

    \end{tabular}
  \endgroup\label{tab:sptcv-methods}
\end{table}

## Spatial

Spatial partitioning methods operate in a two-dimensional space and split observations into training and test set by their spatial location.
Methods such as grid-based partitioning or clustering are used to create groups of observations which may then be aggregated further to form the final partition sets.

Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table&nbsp;\@ref(tab:sptcv-methods)).
If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.

### Buffering

"Buffering" refers to the idea of removing observations located between test and training sets by drawing a circular buffer around one or more observations [@brenning2012; @rest2014; @blockCV].
Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.

Buffering is mostly applied in the coordinate space.
However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).

#### Leave-one-disc-out --- sperrorest

Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set and optionally excludes a buffer zone from the remaining training data.
It thus ensures that a minimum separation distance between training and test data is maintained.
Discs are centered at sample locations, their number being specified by the `folds` argument.
It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
The number of discs is specified via the `folds` argument.

This resampling method is, strictly speaking, not a partitioning method since the discs are not a disjoint.
Nevertheless, it is a logical extension of spatial LOO CV.
Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.

#### Spatial buffering --- blockCV

The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out (LOO) cross-validation.
It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling [@hijmans2013].
However, by default, the method considers all points for sampling the training and test sets.

@@

<!-- TODO: Discuss issues setting a sensible range value -->

### Spatial blocking

Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV].
Several methods have been implemented to create rectangular as well as irregularly shaped blocks. 
These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level.

\pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks: 
(1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks.
In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)).

#### Square and rectangular blocks --- blockCV

Block-level resampling from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`method = "spcv_block"` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
The implementation in \pkg{blockCV} has specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
Users can also supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.

@@

#### Leave-one-tile-out resampling --- sperrorest

#### Leave-one-block-out resampling using coordinate-based clustering --- sperrorest

Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
This can be regarded as a leave-one-block-out resampling method, or as a $k$-fold CV in which each test set is a spatial cluster.
This method is referred to as `spcv_coords` in \pkg{mlr3spatiotempcv}.

Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in Section \@ref(sec:custom-cv)).

## Spatial, temporal or spatiotemporal

### Group-level resampling {#sec:group-level}

@@

#### Leave-Location-And-Time-Out --- CAST

##### Leave-Time-Out (LTO)

##### Leave-Location-Out (LLO)

##### Leave-Location-and-Time-Out (LLTO)

### Custom and grouped resampling --- mlr3 {#sec:custom-cv}

Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
This partitioning concept is also often referred to as "grouped resampling" and should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which inherits additional aggregation of groups at the partition level.
Each level of the factor vector will result in one partition.
Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.

autoplot(resampling_custom, size = 0.8, task = task) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))

```

## Feature space clustering

@@

## Additional partitioning techniques

@@

>>>>

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table&nbsp;\@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
The following sections give an overview of most existing partitioning strategies and their visualization options.
The available methods are further discussed in section \@ref(sec:disc).

\begin{table}[h]
  \centering
  \footnotesize
  \caption[t]{Spatiotemporal resampling methods available in the mlr3 ecosystem.
    Column name shows the mlr3 method name as found in the `mlr3::mlr\_resamplings` dictionary.
    Column "Count" refers to the number of studies which were found to have used this resampling technique until May 2021.
    Literature appearances were only collected for methods with a "spcv" or "sptcv" prefix as methods which rely on custom grouping ("custom\_cv" and "cv") do not have a dedicated name or identifier to make them detectable in research papers.
    For each method, up to three randomly chosen references were added to the table as footnotes.}
  \begingroup
    \begin{tabular}{lllll}
      \\
      Type                            & Name                        & R Package     & Sub-Type  & Count\\
      \toprule
      \multirow{2}{*}{Spatial Leave-one-out}        &  \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}} \tablefootnote{\label{diesing2020}\cite{diesing2020}} & \pkg{blockCV}  & buffer, LOO & 2   \\
      \cmidrule{2-5}
       & \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}} & \pkg{sperrorest}      & buffer, multiple observations  & 3   \\
      \midrule
      \multirow{4}{*}{Leave-one-block-out CV} & \texttt{"spcv\_coords"}  \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} & \pkg{sperrorest}       &  clustering & 6 \\
            \cmidrule{2-5}
                                      & \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} & \pkg{sperrorest}    &  geometric &  4  \\
                                                  \cmidrule{2-5}
                                      &  \texttt{"custom\_cv"}           & \pkg{mlr3} & custom  & - \\
      \midrule
      \multirow{4}{*}{CV at the block level} & \texttt{"spcv\_block"}  \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} & \pkg{blockCV}       &  geometric & 28 \\
  \cmidrule{2-5}
                                      & \texttt{"cv"}  & \pkg{mlr3}    &  custom, honors grouping &  -  \\
  \cmidrule{2-5}
                                      & \texttt{"spcv\_env"} \footref{morera2021}  & \pkg{blockCV}    &  clustering, feature space &  1  \\
      \midrule
      \multirow{3}{*}{CV for spatiotemporal data} & \texttt{"sptcv\_cstf"}  \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & \pkg{CAST}       &  custom & 6 \\
                                                  \cmidrule{2-5}
                                      & \texttt{"sptcv\_cluto"}  & \pkg{skmeans}    &  clustering & - \\
                                      \bottomrule
    \end{tabular}
  \endgroup\label{tab:sptcv-methods}
\end{table}

On the most upper level, resampling methods are categorized through the way the fold partitioning is done:

- Leaving out single pixels (Spatial Leave-one-out) per fold with an optional buffer
- Leaving out single blocks (Leave-one-block-out CV) per fold with an optional buffer
- Leaving out multiple blocks (CV at the block level) per fold

A finer distinction can then be made by looking at how these groups are formed by the respective individual method:

- Using a geometric-based approach (rectangular or circular)
- Using an unsupervised clustering approach
- Using a custom/external input, i.e. specifying the groups via an external grouping variable

The terminology "block" used in this categorization and in the manuscript in general refers to any kind of regular and irregularly shaped grouping of observations and not necessarily only to a rectangular partitioning as used in some methods.

<!-- explain and introduce buffering -->
The use of a buffer is not a direct grouping characteristic of a partitioning method in our view but only an optional add-on which could potentially be used by any method.
"Buffering" refers to the idea of removing observations located between test and training sets by drawing a circular buffer around one or more observations [@brenning2012; @rest2014; @blockCV].
Within this buffer zone, all observations are removed from the training set, thereby reducing the similarity between training and test set.
Buffering is mostly applied in the coordinate space.
However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) `"spcv\_buffer"` from package \pkg{blockCV} and (2) `"spcv\_disc"` from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (Table&nbsp;\@ref(tab:sptcv-methods)).

<!-- FIXME: graphic here -->

## Spatial Leave-one-out

Spatial Leave-one-out methods start from a single observation in space and create radial (buffer) zones around which define the partitions.

### Spatial Leave-one-out with buffer --- `"spcv_buffer"`

The "buffering" method from the \pkg{blockCV} package is conceptually related to spatial leave-one-out cross-validation.
It comes with optional support for presence/absence data handling, as often used in the field of species distribution modeling [@hijmans2013].
However, by default, the method considers all points for sampling the training and test sets.

@@

### Leave-one-disc-out with optional buffer --- `"spcv_disc"`

Leave-one-disc-out resampling from package \pkg{sperrorest} randomly selects a circular test set starting from a single observations and optionally excludes a buffer zone from the remaining training data.
It thus ensures that a minimum separation distance between training and test data is maintained.
Discs are centered at sample locations, their number being specified by the `folds` argument.
It is therefore possible to sample up to $n$ discs without replacement (default) or any number when sampling with replacement (`replace = TRUE`).
The number of discs is specified via the `folds` argument.

This resampling method is, strictly speaking, not a partitioning method since the discs are not resulting in disjoint partitions - partitions will potentially overlap in subsequent folds.
Nevertheless, it is a logical extension of spatial LOO CV.
Depending on the size of the discs, potentially a large number of discs (`folds` argument) is needed to effectively use the entire dataset.

Even though the buffer creates a circular area which defines the test partition, it is not categorized as a "geometric" method as the origin of the circle is observation-based.
In contrast, "geometric" methods do not take into account single observation locations but only focus on the study area extend and will include all observations which fall into their respective covered area.

## Leave-one-block-out Cross-Validation

Leave-one-block-out CV methods partition the dataset in a way that each of the resulting partitions will be used as a fold in the CV.

### Clustering-based: Using coordinate-based clustering - `"spcv_coords"`

Coordinate-based clustering for spatial CV [@russ2010; @brenning2012] as implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.
This can be regarded as a leave-one-block-out resampling method, or as a $k$-fold CV in which each test set is a spatial cluster.
This method is referred to as `"spcv_coords"` in \pkg{mlr3spatiotempcv}.

Coordinate-based clustering approach is very versatile as it adapts to irregularly-shaped study areas and ensures that exactly $k$ partitions are created, which are usually or very similar size when the sample locations are spread out evenly.
Nevertheless, despite the random selection of initial cluster centers, repeated partitionings may in some cases be nearly identical.
Also, $k$-means clustering may be less suitable for data sets with pre-existing clusters of points and/or with distant isolated sample locations.
When distinct clusters of points are present, as in multi-level sampling, it may be better to define clusters using a factor variable (see method `"custom_cv"` in section \@ref(sec:custom-cv)).

### Geometric: using rectangular blocks --- `"spcv_tiles"`

### Custom: `"custom_cv"` in mlr3

Support for user-defined partitioning strategies is built into \pkg{mlr3} directly.
In this so-called "Custom CV", users are able to supply an external factor vector with the same length as number of observations.
This partitioning concept should not be mistaken with "group-level resampling" (section \@ref(sec:group-level)), which conducts additional combination of data splits and performs CV at the block level.
Each level of the factor vector will result in one partition.
Alternatively to supplying an external factor vector, the name of a feature (type `character`) present within the task can be supplied via argument `col`.

autoplot(resampling_custom, size = 0.8, task = task) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

## Cross-Validation at the block level

Methods which operate on the block level create dataset splits which are then combined to form the final partitions for the CV folds.
In the special case of an equal amount of dataset splits and requested CV folds, these methods would perform a Leave-one-block-out CV.

### Geometric: using rectangular blocks --- `"spcv_block"`

The `"spcv_block"` method from package \pkg{blockCV} supports both random and systematic resampling of square blocks (`rsmp("spcv_block")` with argument `selection = "random"` and `"systematic"`, respectively; Figure&nbpsp;~\@ref(fig:block-random) and \@ref(fig:block-systematic)).
This implementation comes with specific options that are designed for modeling presence-only data, which is a typical situation in species distribution modeling.
Users can also supply a user-defined polygon via argument `rasterLayer` with predefined blocking zones.

@@

### Custom: `"cv"` in mlr3 {#sec:group-level}

@@

### Clustering: using feature-based clustering --- `"spcv_env"`

@@

## Cross-Validation for spatiotemporal data

### Custom: "Leave-location-and-time-out" --- `"spcv_cstf"`

The sub-methods "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO) are conceptually similar to the "cv" method with a custom grouping from \pkg{mlr3} as they perform a CV at the block level using a predefined grouping structure.

#### Leave-Time-Out (LTO)

#### Leave-Location-Out (LLO)

#### Leave-Location-and-Time-Out (LLTO)

### Clustering: using CLUTO --- `"sptcv_cluto"`

@@

<!-- Spatial blocking divides the study area into different zones [russ2010; @bahn2012; @wenger2012; @blockCV]. -->
<!-- Several methods have been implemented to create rectangular as well as irregularly shaped blocks.  -->
<!-- These blocks can be used for leave-one-block-out resampling, or they can themselves be grouped into $k$ partitions for a $k$-fold CV at the block level. -->

<!-- \pkg{mlr3spatiotempcv} supports multiple resampling methods that are based on blocks:  -->
<!-- (1) "Spatial blocking" from package \pkg{blockCV} implements $k$-fold CV resampling at the level of square or rectangular blocks, while (2) "Spatial tiles" from \pkg{sperrorest} offers leave-one-block-out resampling with square or rectangular, possibly rotated blocks. -->
<!-- In addition, irregularly shaped blocks are available (3) through coordinate-based clustering as a leave-one-block-out technique from \pkg{sperrorest} (method `"spcv_coords"`, Section&nbsp;\@ref(sec:cluster-cv)), and (4) as a CV at the block level where blocks are defined by a factor variable that defines zones or groups (method `"custom_cv"` in \pkg{mlr3}, Section&nbsp;\@ref(sec:custom-cv)). -->
